Balance responses in MS

### 2020-02-04 R/msBalance.Rmd
added notes about normalization. probably better to average across replicates prior to calculating max values, will be less sensitive to high frequency data.
additions to plot normalization

```{r loadData}
library(vroom)
library(dplyr)
library(magrittr)
library(stringr)
library(lubridate)
`%contains%` = function(s, pattern) grepl(pattern, s)
`%omits%` = function(s, pattern) !grepl(pattern, s)

@@

```{r layout}
# layout tables and figures
library(captioner)
tableNumbers <- captioner(prefix = "Table")
captionDemographics = tableNumbers(name = "demographics", caption = "Clinical and demographic features of the study population.")
figureNumbers <- captioner(prefix = "Figure")
captionCartoon = figureNumbers(name = "cartoon", caption = "Perturbation directions used in testing protocol.")
captionRawEmg = figureNumbers(name = "rawEmg", caption = "EMG traces stratified by perturbation direction and study group.")
```

@@

## Data processing
Each individual EMG record was registered to the onset of treadmill motion and resampled to a common sample rate (1 kHz), and processed (high-pass, 35 Hz, de- mean, rectify, low-pass, 40 Hz) as in previous studies.[@lang2019; @mckay2016]

@@

# I cannot figure out how to get dplyr::left_join to join by a variable with backticks, so rename temporarily
taR = left_join(taR, trialData, by = "Code") %>% 
  select(`Study Group`,Participant, Date, Trial, `Code`, `Trial Number`, `Perturbation Direction`, Muscle, Side, everything()) %>% 
  arrange(`Study Group`, Participant, `Trial Number`, Time)

# note that a large amount of the data (40%) is follow-up, with different visit amounts. for now, let's just consider the initial visit.
taR = taR %>% filter(Visit == "Initial")

# consider only those participants for whom both some EMG and minimal demographic data are available.
includedParticipants = intersect(taR$Participant, participants$Participant)
taR = taR %>% filter(Participant %in% includedParticipants)
participants = participants %>% filter(Participant %in% includedParticipants)

taR = taR %>% mutate(Group = factor(`Study Group`, levels = c("Control", "MS"))) %>% select(-`Study Group`) %>% 
  mutate(`Perturbation Direction` = factor(`Perturbation Direction`, levels = c("Forward", "Backward")))

# calculate max activity for normalization purposes. this has already been filtered, etc. so it is probably ok just to normalize to maximum values.
taRNorm = taR %>% group_by(Participant) %>% summarize(Vm = max(V, na.rm = TRUE)) %>% ungroup()
taR = taR %>% left_join(taRNorm, by = "Participant") %>% mutate(Vn = V/Vm)

```

@@

***
```{r demographics}
library(janitor)
library(kableExtra)
participants %>% tabyl(Sex, Group) %>% mutate(Total = MS + Control) %>% 
  kable(caption = captionDemographics) %>% kable_styling(full_width = F, position = "left")
```
***

## Changes in tibialis anterior activity with disease state and perturbation direction
Recorded EMG traces are presented in `r figureNumbers("rawEmg", display = "cite")`.

***
```{r rawEMG, out.width = "100%", fig.align = "left", fig.cap = captionRawEmg}
library(ggplot2)
library(cowplot)
p = ggplot(taR, aes(x = Time, y = Vn, group = Code)) + geom_line() + facet_grid(Group ~ `Perturbation Direction`) + theme_minimal_grid()
p
```
***

>>>>

```{r loadData}
library(vroom)
library(dplyr)
library(magrittr)
library(stringr)
library(lubridate)
library(janitor)
library(kableExtra)
library(ggplot2)
library(cowplot)
`%contains%` = function(s, pattern) grepl(pattern, s)
`%omits%` = function(s, pattern) !grepl(pattern, s)
`%between%` = function(s, window) s>=min(window)&s<max(window)

@@

```{r layout}
# layout tables and figures
library(captioner)
tableNumbers <- captioner(prefix = "Table")
captionEmgMags = tableNumbers(name = "emgMags", caption = "Maximum EMG levels.")
captionDemographics = tableNumbers(name = "demographics", caption = "Clinical and demographic features of the study population.")

figureNumbers <- captioner(prefix = "Figure")
captionCartoon = figureNumbers(name = "cartoon", caption = "Perturbation directions used in testing protocol.")
captionRawEmg = figureNumbers(name = "rawEmg", caption = "EMG traces stratified by perturbation direction and study group, prior to normalization. Colors represent individual participants.")
```

@@

## Data processing
Each individual EMG record was registered to the onset of treadmill motion and resampled to a common sample rate (1 kHz), and processed (high-pass, 35 Hz, de- mean, rectify, low-pass, 40 Hz) as in previous studies.[@lang2019; @mckay2016]
Recorded EMG records were visualized stratified by disease state and perturbation direction (`r figureNumbers("rawEmg", display = "cite")`).
By visual inspection, there were some gross magnitude differences in recorded EMG levels between the MS and control groups, with higher EMG levels overall in the MS group.
EMG levels were therefore normalized to the maximum values obtained in a 1 second window after perturbation onset after averaging across like trials (`r tableNumbers("emgMags", display = "cite")`).

@@

# I cannot figure out how to get dplyr::left_join to join by a variable with backticks, so rename temporarily
taR %<>% left_join(trialData, by = "Code") %>% 
  select(`Study Group`,Participant, Date, Trial, `Code`, `Trial Number`, `Perturbation Direction`, Muscle, Side, everything()) %>% 
  arrange(`Study Group`, Participant, `Trial Number`, Time)

# note that a large amount of the data (40%) is follow-up, with different visit amounts. for now, let's just consider the initial visit.
taR %<>% filter(Visit == "Initial")

# consider only those participants for whom both some EMG and minimal demographic data are available.
includedParticipants = intersect(taR$Participant, participants$Participant)

# consider only a subset of participants if debugging
SP = FALSE
if (SP){
  includedParticipants = sample(includedParticipants, 60)
}

taR %<>% filter(Participant %in% includedParticipants)
participants %<>% filter(Participant %in% includedParticipants)

# here are the time records
taR %<>% mutate(Group = factor(`Study Group`, levels = c("Control", "MS"))) %>% select(-`Study Group`) %>% 
  mutate(`Perturbation Direction` = factor(`Perturbation Direction`, levels = c("Forward", "Backward")))

# here are the averages across like trials
taRAv = taR %>% group_by(Group, Participant, Muscle, Side, `Perturbation Direction`, Time) %>% summarize(V = mean(V, na.rm = TRUE)) %>% ungroup()

# to calculate normalization coefficients, average across like trials for each patient, limit to a particular normalization window, and calculate maximum values
taRNorm = taRAv %>% filter(Time %between% c(0.0, 1.0)) %>% group_by(Group, Participant, `Perturbation Direction`) %>% summarize(Vm = max(V, na.rm = TRUE)) %>% ungroup()
```

***
```{r rawEMG, out.width = "100%", fig.align = "left", fig.cap = captionRawEmg}
p = ggplot(taR, aes(x = Time, y = V)) + geom_line(aes(color = Participant), show.legend = FALSE, alpha = 0.5) + facet_grid(Group ~ `Perturbation Direction`) + theme_minimal_grid() +
  geom_jitter(mapping = aes(x = 0, y = Vm, color = Participant), data = taRNorm %>% left_join(participants %>% select(Participant, Group)), show.legend = FALSE, height = 0, width = 0.01)
p
```

```{r emgMags}
emgMags = 
  taRNorm %>% group_by(`Perturbation Direction`, Group) %>% summarize(Mean = mean(Vm) %>% round() ) %>% left_join(
    taRNorm %>% group_by(`Perturbation Direction`, Group) %>% summarize(SD = sd(Vm) %>% round() ) ) %>% left_join(
      taRNorm %>% group_by(`Perturbation Direction`, Group) %>% summarize(Max = max(Vm) %>% round() ) )
emgMags %>% kable(caption = captionEmgMags) %>% kable_styling(full_width = F, position = "left")

# 
# 
# ggplot(taRAv, aes(Time, Vn, color = Participant)) + geom_point(show.legend = FALSE) + facet_wrap(vars(`Perturbation Direction`))
# 
# p = ggplot(taRAv, aes(Time, Vn)) + geom_point()
# 
# p = ggplot(taRAv %>% group_by(Participant, `Perturbation Direction`), aes(x = Time, y = Vn, color = Participant)) +
#   facet_grid(Group ~ `Perturbation Direction`) +
#   theme_minimal_grid() +
#   coord_cartesian(xlim = c(0,1)) + 
#   geom_line(aes(group = Participant, color = Participant), taRav, show.legend = FALSE) +
# p
```

***

@@

***
```{r demographics}
participants %>% tabyl(Sex, Group) %>% mutate(Total = MS + Control) %>% 
  kable(caption = captionDemographics) %>% kable_styling(full_width = F, position = "left")
```
***

## Changes in tibialis anterior activity with disease state and perturbation direction
